---
title: Confidence Intervals
date: today
date-format: long
footer: "[IAFF 6501 Website](https://quant4ia.rocks/)"
logo: images/iaff6501-logo.png
format:
  revealjs:
    theme: [simple, custom.scss]
    transition: fade
    slide-number: true
    #multiplex: true
    chalkboard: true
execute:
  echo: false
  message: false
  warning: false
  freeze: auto
---

## {.smaller}

On December 19, 2014, the front page of Spanish national newspaper El
País read *"Catalan public opinion swings toward 'no' for independence, says survey"*.^[Alberto Cairo. [The truthful art: Data, charts, and maps for communication](http://www.thefunctionalart.com/p/the-truthful-art-book.html). New Riders, 2016.]

```{r}
library(tidyverse)
library(lubridate)
library(scales)
catalan <- tibble(
  response = c("No", "Yes", "No answer"),
  rate     = c(45.3, 44.5, 10.2)
) %>%
  mutate(response = fct_relevel(response, "No", "Yes", "No answer"))
```

```{r}
#| label: catalan-misleading
ggplot(catalan, aes(y = fct_rev(response), x = rate, fill = response)) +
  geom_col(width = .75) +
  scale_fill_manual(values = c("#5C8AA9", "#9D303A", "gray")) +
  scale_x_continuous(labels = label_percent(scale = 1)) +
  theme_minimal(base_size = 16) +
  labs(
    title = "Do you want Catalonia\nto become an independent state?",
    caption = "Margin of error: +/-2.95% at 95% confidence level",
    x = NULL, y = NULL) +
  theme(legend.position = "none")
```

## {.smaller}

The probability of the tiny difference between the 'No' and 'Yes' being just due to random chance is very high.^[Alberto Cairo. ["Uncertainty and Graphicacy"](https://ec.europa.eu/eurostat/cros/powerfromstatistics/OR/PfS-OutlookReport-Cairo.pdf), 2017.]

```{r}
catalan <- catalan %>%
  mutate(
    low = rate - 2.95,
    high = rate + 2.95
  )
ggplot(catalan, aes(y = fct_rev(response), x = rate, color = response, group = response)) +
  geom_segment(aes(x = low, xend = high, 
                   y = fct_rev(response), yend = fct_rev(response)),
               linewidth = 0.8, color = "black") +
  geom_point(size = 3) +
  scale_color_manual(values = c("#5C8AA9", "#9D303A", "gray")) +
  scale_x_continuous(labels = label_percent(scale = 1)) +
  guides(color = "none") +
  theme_minimal(base_size = 16) +
  labs(
    title = "Do you want Catalonia to become an independent state?",
    x = NULL, y = NULL
  )
```

## Characterizing Uncertainty

<br>

- We know from previous section that even unbiased procedures do not get the "right" answer every time
- We also know that our estimates might vary from sample to sample due to random chance
- Therefore we want to report on our estimate and our level of uncertainty


## Characterizing Uncertainty

<br>

- With M&Ms, we knew the population parameter
- In real life, we do not!
- We want to generate an estimate *and* characterize our uncertainty with a *range* of possible estimates
    
## Solution: Create a Confidence Interval

<br>

- A plausible range of values for the population parameter is a confidence interval.

. . .

- 95 percent confidence interval is standard
    - We are 95% confident that the parameter value falls within the range given by the confidence interval

## Ways to Estimate

<br>

- Take advantage of Central Limit Theorem to estimate using math
- Use simulation, bootstrapping 

## With Math...

$$CI = \bar{x} \pm Z \left( \frac{\sigma}{\sqrt{n}} \right)$$

- $\bar{x}$ is the sample mean,
- $Z$ is the Z-score corresponding to the desired level of confidence
- $\sigma$ is the population standard deviation, and 
- $n$ is the sample size

## 

<br>

This part here represents the standard error: 

$$\left( \frac{\sigma}{\sqrt{n}} \right)$$

- Standard deviation of the sampling distribution
- Characterizes the spread of the sampling distribution
- The bigger this is the bigger the CIs are going to be

## Central Limit Theorem

$$CI = \bar{x} \pm Z \left( \frac{\sigma}{\sqrt{n}} \right)$$

- This way of doing things depends on the Central Limit Theorem
- As sample size gets bigger, the spread of the sampling distribution gets narrower
- The shape of the sampling distributions becomes more normally distributed

## 

<br>

$$CI = \bar{x} \pm Z \left( \frac{\sigma}{\sqrt{n}} \right)$$

This is therefore a **parametric** method of calculating the CI. It depends on assumptions about the normality of the distribution.  

## Bootstrapping

<br>

::: incremental
- Pulling oneself up from their bootstraps ... 
- Use the data we have to estimate the sampling distribution
- We call this the *bootstrap* distribution
- This is a **nonparametric** method
- It does not depend on assumptions about normality
:::

## Bootstrap Process {.smaller}

<br>

1. Take a bootstrap sample - a random sample taken **with replacement** from the original sample, of the **same size** as the original sample

. . .

2. Calculate the bootstrap statistic - a statistic such as mean, median, proportion, slope, etc. computed on the bootstrap samples

. . .

3. Repeat steps (1) and (2) many times to create a bootstrap distribution - a distribution of bootstrap statistics

. . .

4. Calculate the bounds of the XX% confidence interval as the middle XX% 
of the bootstrap distribution (usually 95 percent confidence interval)


## Russia

<br>

What Proportion of Russians believe their country interfered in the 2016 presidential elections in the US?

- Pew Research survey
- 506 subjects
- Data available in the `openintro` package

## 

<br>

For this example, we will use data from the Open Intro package. Install that package before running this code chunk.

<br>

```{r}
#| echo: true

#install.packages("openintro")
library(openintro)

glimpse(russian_influence_on_us_election_2016)
```


## 

<br>

Let's use `mutate()` to recode the qualitative variable as a numeric one...

<br>

```{r}
#| echo: true

russiaData <- russian_influence_on_us_election_2016 |> 
  mutate(try_influence = ifelse(influence_2016 == "Did try", 1, 0))
```

##

<br>

Now let's calculate the mean and standard deviation of the `try_influence` variable... 

<br>

```{r}
#| echo: true

russiaData |>
  summarize( 
          mean = mean(try_influence),
          sd = sd(try_influence)
  )
```

## 

<br>

And finally let's draw a bar plot...

<br>

```{r}
#| echo: true
#| eval: false

ggplot(russiaData, aes(x = try_influence)) +
  geom_bar(fill = "steelblue", width = .75) +
  labs(
    title = "Did Russia try to influence the U.S. election?",
    x = "0 = 'No', 1 = 'Yes'",
    y = "Frequncy"
  ) +
  theme_minimal()
```

## 

<br>

```{r}
ggplot(russiaData, aes(x = try_influence)) +
  geom_bar(fill = "steelblue", width = .75) +
  labs(
    title = "Did Russia try to influence the U.S. election?",
    x = "0 = 'No', 1 = 'Yes'",
    y = "Frequncy"
  ) +
  theme_minimal()
```

## Bootstrap with `tidymodels`

Install `tidymodels` before running this code chunk... 

```{r}
#| echo: true

#install.packages("tidymodels")
library(tidymodels)

set.seed(66)
boot_df <- russiaData |>
  # specify the variable of interest
  specify(response = try_influence) |>
  # generate 15000 bootstrap samples
  generate(reps = 15000, type = "bootstrap") |>
  # calculate the mean of each bootstrap sample
  calculate(stat = "mean")

glimpse(boot_df)
```

## 

<br>

Calculate the confidence interval. A 95% confidence interval is bounded by the middle 95% of the bootstrap distribution.

<br>

```{r}
#| echo: true 

boot_df |>
  summarize(lower = quantile(stat, 0.025),
            upper = quantile(stat, 0.975))
```

## 

<br>

Create upper and lower bounds for visualization.

<br>

```{r}
#| echo: true

# for using these values later
lower_bound <- boot_df |> summarize(lower_bound = quantile(stat, 0.025)) |> pull() 
upper_bound <- boot_df |> summarize(upper_bound = quantile(stat, 0.975)) |> pull() 
```

##

<br>

Visualize with a histogram

<br>

```{r}
#| echo: true
#| eval: false

ggplot(data = boot_df, mapping = aes(x = stat)) +
  geom_histogram(binwidth =.01, fill = "steelblue4") +
  geom_vline(xintercept = c(lower_bound, upper_bound), color = "darkgrey", size = 1, linetype = "dashed") +
  labs(title = "Bootstrap distribution of means",
       subtitle = "and 95% confidence interval",
       x = "Estimate",
       y = "Frequency") +
  theme_bw()
```

## 

<br>

```{r}
ggplot(data = boot_df, mapping = aes(x = stat)) +
  geom_histogram(binwidth =.01, fill = "steelblue4") +
  geom_vline(xintercept = c(lower_bound, upper_bound), color = "darkgrey", size = 1, linetype = "dashed") +
  labs(title = "Bootstrap distribution of means",
       subtitle = "and 95% confidence interval",
       x = "Estimate",
       y = "Frequency") +
  theme_bw()
```

## Interpret the confidence interval {.smaller}

<br>

The 95% confidence interval was calculated as (`lower_bound`, `upper_bound`). Which of the following is the correct interpretation of this interval?

<br>

**(a)** 95% of the time the percentage of Russian who believe that Russia interfered in the 2016 US elections is between `lower_bound` and `upper_bound`.

**(b)** 95% of all Russians believe that the chance Russia interfered in the 2016 US elections is between `lower_bound` and `upper_bound`.

**(c)** We are 95% confident that the proportion of Russians who believe that Russia interfered in the 2016 US election is between `lower_bound` and `upper_bound`.

**(d)** We are 95% confident that the proportion of Russians who supported interfering in the 2016 US elections is between `lower_bound` and `upper_bound`.

## Your Turn! {.smaller}

<br>

- Change the `reps` argument in the `generate()` function to 1000. What happens to the width of the confidence interval?
- Change the `reps` argument in the `generate()` function to 5000. What happens to the width of the confidence interval?
- Change the `reps` argument in the `generate()` function to 10000. What happens to the width of the confidence interval?
- How does the width of the confidence interval change as the number of bootstrap samples increases?
- How would you interpret this finding? 

## Bias vs Precision

- A procedure is *unbiased* if it generates the "right" answer, on average
- Precision refers to variability: procedures with less sampling variability will be more precise
    - all else equal, a greater sample size will increase precision
- When we increase the sample size (number of reps), we increase precision
- As a result our confidence interval will be narrower

## Why did we do these simulations?

<br>

- They provide a foundation for statistical inference and for characterizing *uncertainty* in our estimates
- The best **research designs** will try to maximize or achieve good balance on bias vs precision





