---
title: Sampling
date: today
date-format: long
footer: "[IAFF 6501 Website](https://quant4ia.rocks/)"
logo: images/iaff6501-logo.png
format:
  revealjs:
    theme: [simple, custom.scss]
    transition: fade
    slide-number: true
    #multiplex: true
    chalkboard: true
execute:
  echo: false
  message: false
  warning: false
  freeze: auto
---

## Sampling

::: incremental
- **Sampling** the act of selecting a subset of individuals, items, or data points from a larger population to estimate characteristics or metrics of the entire population
- Versus a **census**, which involves gathering information on every individual in the population
- Why would you want to use a sample?
:::

Today's code is [HERE](https://www.dropbox.com/scl/fo/3u8lgap6nqtcp91gmclgd/h?rlkey=wm2ey4emths9w9585gnp9fygb&dl=0)

# Sampling Activity

## What proportion of all milk chocolate M&Ms are blue?

-   M&Ms has a precise distribution of colors that it produces in its factories
-   M&Ms are sorted into bags in factories in a fairly random process
-   Each bag represents a **sample** from the full **population** of M&Ms

## Activity

-   Get in groups of 2. Each group will have 4-5 bags of M&Ms.
-   Keep the contents of each bag separate, and do not eat (yet!)
-   Open up your first bag of M&Ms: calculate the proportion of the M&Ms that are blue. **Write this down.** What is your best guess (your **estimate**) for the proportion of all M&Ms that are blue?

## Activity

-   Do the same as above for the rest of your bags (you should have 4-5 **estimates** written down)
-   Draw a histogram of your estimates (by hand)
-   Add your estimates to this [Google Sheet](https://docs.google.com/spreadsheets/d/136wGKZOnwOdo3O-4bUfF_TWKBCAWQUCyiVKSY4HyYAw/edit?usp=sharing)
-   Add your estimates to the class histogram on the whiteboard

## Let's Analyze the Data

<br>

We will use the `googlesheets4` package to pull the data into R so be sure to install it.

```{r}
#| label: load_mm_data
#| echo: true
#| eval: false

#install.packages("googlesheets4")

library(tidyverse)
library(googlesheets4)

gs4_deauth() # to signify no authorization required

mnm_data <- read_sheet("https://docs.google.com/spreadsheets/d/136wGKZOnwOdo3O-4bUfF_TWKBCAWQUCyiVKSY4HyYAw/edit#gid=0")

glimpse(mnm_data)
```


## Calculate Some Summary Stats

<br>

```{r}
#| label: summary_stats
#| echo: true
#| eval: false

mnm_data |>
  summarize(
    mean_blue = mean(proportion_blue),
    median_blue = median(proportion_blue),
    sd_blue = sd(proportion_blue)
  )
```
## Now Let's Make a Histogram 

<br>

```{r}
#| label: histogram
#| echo: true
#| eval: false

ggplot(mnm_data, aes(x = proportion_blue)) +
  geom_histogram(fill = "steelblue") +
  labs(
    title = "Percentage of Blue M&Ms",
    x = "Proportion Blue",
    y = "Count"
  )
```

## Discuss with Neighbor 

-   What is the histogram/distribution showing?
-   Based on the histogram on the board, what is your answer to the question of what proportion of all milk chocolate M&Ms are blue? Why do you give that answer?
-   Why do some bag of M&Ms have proportions of blues that are higher and lower than the number you gave above?
-   How do our estimates relate to the actual percentage of blue M&Ms manufactured (ask Google or ChatGPT)

## What did we just do?

::: incremental
-   We wanted to say something about the **population** of M&Ms
-   The **parameter** we care about is the proportion of M&Ms that are blue
-   It would be impossible to conduct a **census** and to calculate the parameter
-   We took a **sample** from the population and calculated a **sample statistic**
-   **statistical inference:** act of making a guess about a **population** using information from a **sample**
:::

## What have we just done?

::: incremental
-   We completed this task many times
-   This produced a **sampling distribution** of our **estimates**
-   There is a distribution of estimates because of **sampling variability** 
-   Due to random chance, one estimate from one sample can differ from another
-   These are foundational ideas for statistical inference that we are going to keep building on
:::

# Zooming Out

## Target Population

In data analysis, we are usually interested in saying something about a **Target Population.**

::: incremental
-   What proportion of adult Russians support the war in Ukraine?
    -   Target population: adult Russians (age 18+)
-   How many US college students check social media during their classes?
    -   Target population: US college students
-   What percentage of M&Ms are blue?
    -   Target population: all of the M&Ms
:::

## Sample

<br>

In many instances, we have a **Sample**

-   We cannot talk to every Russian
-   We cannot talk to all college students
-   We cannot count all of the M&Ms

## Parameters vs Statistics

<br>

::: incremental
-   The **parameter** is the value of a calculation for the entire target population
-   The **statistic** is what we calculate on our sample
    -   We calculate a statistic in order to say something about the parameter
:::

## Inference

<br>

::: incremental
-   **Inference**--The act of "making a guess" about some unknown
-   **Statistical inference**--Making a good guess about a population from a sample
-   **Causal inference**--Did X cause Y? [topic for later classes]
:::

